|

### 2022-07-23 manuscript.Rmd
some house-cleaning

---
title: |
  Teaching Visual Accessibility in the Introductory Data Science Classes: Why, What, When, and How
shorttitle: Teaching Accessibility in Data Science
author:
- name: JooYoung Seo
  affiliation: "1"  # Must be quoted when including comma.
  email: "jseo1005@illinois.edu"  # Only corresponding authors require email.
- name: Mine Dogucu
  affiliation: "2,3"

@@

abstract: |
  Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency. Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and sometimes people with learning disabilities. We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation. In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility. As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
keywords: 
  - Data representations
  - Curriculum
  - R

@@

bibliography: JDSbib.bib
output:
  rticles::jds_article: default
  bookdown::pdf_book:
    base_format: rticles::jds_article # for using bookdown features like \@ref()
---

```{r setup, include = FALSE, echo = FALSE, purl = FALSE}
knitr::opts_chunk$set(echo = FALSE)

@@

# Introduction {#sec:intro}

According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula and a consensus on what should be included in such curricula has not yet been reached. 
In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula \cite{danyluk2021computing}.

@@

In this proposed talk, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
We teach R as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.

@@

# Learning Goals {#sec:goals}

@@

Currently, `BrailleR::VI()` function has a capability to auto-generate alt text for basic R graphics and ggplot objects. 
Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.

@@

Our definition of sonification refers to a data representation method using stereo sound and various audible patterns. For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)`. 
In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.

@@

Data tactulization refers to making data visualization in a tactile form so that it can be touchable by fingers. 
Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired. 
One of the authors in this talk, developed a simple R package, called `tactileR` that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()`.

@@

>>>>

---
title: |
  Teaching Visual Accessibility in the Introductory Data Science Classes: Why, What, When, and How
shorttitle: Teaching Accessibility in Data Science # nolint # nolint
author:
- name: JooYoung Seo
  affiliation: "1"  # Must be quoted when including comma.
  email: "jseo1005@illinois.edu"  # Only corresponding authors require email.
- name: Mine Dogucu
  affiliation: "2,3"

@@

abstract: |
  Although there are various ways to represent data patterns and models, visualization has been primarily taught in many data science courses for its efficiency. Such vision-dependent output may cause critical barriers against those who are blind and visually impaired, and people with learning disabilities. We argue that instructors need to teach multiple data representation methods so that all students of all abilities can benefit from multi-sensory data augmentation. In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility. As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
keywords: 
  - Data representations
  - Curriculum
  - R

@@

bibliography: ["bib/references.bib", "bib/packages.bib"]
output:
  rticles::jds_article: default
  bookdown::pdf_book:
    base_format: rticles::jds_article # for using bookdown features like \@ref()
---

```{r setup, include=FALSE, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = FALSE)

@@

```{r pkg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, purl=FALSE}
# Load required packages
library(ggplot2)
library(scales)
library(BrailleR)
library(sonify)
library(tactileR)
library(tidyverse)
```

# Introduction {#intro}

According to LinkedIn's 2017 U.S. Emerging Jobs Report, machine learning engineers and data scientists rank among the top emerging jobs.
With the urgent need for training a higher number of skilled data scientists, many institutes of higher education are developing their own data science curricula, and a consensus on what should be included in such curricula has not yet been reached. 
In the latest version of *Computing Competencies for Undergraduate Data Science Curricula* by ACM Data Science Task Force, accessibility is a recommended content for the data science curricula [@danyluk2021computing].

@@

In this paper, we will argue that accessibility should be taught as early as the introductory course as part of the data science curriculum so that regardless of whether learners major in data science or not they can have foundational exposure to accessibility.
As data science educators who teach accessibility as part of our lower-division courses in two different institutions, we will share specific examples as well as open-access resources that can be utilized by other data science instructors.
We teach R [@R-base] as the main language in our courses thus our examples would be shaped around using R, however, instructors using a different set of language(s) or even teaching language-agnostic courses may still find the content applicable for their own courses.

@@

# Learning Goals {#goals}

@@

Currently, `BrailleR::VI()` [@R-BrailleR] function has a capability to auto-generate alt text for basic R graphics and ggplot objects. 
Supported graphic types include as follows: `graphics::hist()`, `graphics::boxplot()`, `ggplot2::geom_bar()`, `ggplot2::geom_histogram()`, and `ggplot2::geom_boxplot()`.

@@

Our definition of sonification refers to a data representation method using stereo sound and various audible patterns. For example, `plot(x = 1:5, y = 1:5)` can be sonified using `sonify::sonify(x = 1:5, y = 1:5)` [@R-sonify]. 
In this example, a scatter plot is represented by a sound. Position on the X axis is communicated using the left-to-right stereo effect; the pitch of the sound indicates position on the Y axis.

@@

Data tactulization refers to making data visualization in a tactile form so that it can be touchable by fingers. 
Although this requires some expensive embossing hardware (e.g., braille embosser or Swell Form Heating Machine), this may be one of the most effective and accessible data representations for those who are blind or visually impaired. 
One of the authors in this paper, developed a simple R package, called `tactileR` [@R-tactileR] that can generate a ready-to-emboss pdf file from R graphics and ggplot objects. 
Students and instructors can produce their own tactile graphics by using the following functions: `tactileR::brl_begin()`; [their graphic object]; `tactileR::brl_end()`.

@@

```{r include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
knitr::write_bib(file = "bib/packages.bib")
```
